Analysing the effects of COVID-19 on Crime in the United States

### 2022-05-01 outputs/paper/paper.Rmd
Included information about methods and also a detailed discussion on the results and limitations. Fixed Minor issues as well.

The key motivation for our work is previous work showing that certain crimes like domestic violence saw a significantly smaller drop as compared to other crimes after the onset of the pandemic [@bullinger2020covid], indicating the importance of the nature of the actual crime committed. Our focus on individual city data is due to known data collection issues for such research [@boman2020has] and is motivated by preliminary studies showing that the crime rates for specific crimes are largely dependent on the city [@ashby2020initial]. There are also other factors such as the George Floyd protests in June 2020 that we hypothesize may have resulted in a noticeable increase in the vandalism and theft crimes in certain cities.

Overall, our paper is a small contribution building upon and motivated by several recent papers that have covered the impact of the pandemic on the human psyche and crime rates across different states and countries [@boman2021global, @campedelli2020disentangling, @nivette2021global] alongside work prompting further research in the area [@eisner2020violence]. We achieve our goal of analyzing the pandemic data in conjunction with crime data obtained from the official city websites, using R statistical programming [@citeR], through time series modeling and also exploratory data analysis.

@@

@@

Vector autoregression (VAR) is a statistical model used to capture the relationship between multiple quantities as they change over time. VAR is a type of stochastic process model. VAR models generalize the single-variable (univariate) autoregressive model by allowing for multivariate time series. VAR models are often used in economics and the natural sciences.
A pth-order VAR is denoted "VAR(p)" and sometimes called "a VAR with p lags". A pth-order VAR model is written as

$${\displaystyle y_{t}=c+A_{1}y_{t-1}+A_{2}y_{t-2}+\cdots +A_{p}y_{t-p}+e_{t},\,}y_{t}=c+A_{1}y_{{t-1}}+A_{2}y_{{t-2}}+\cdots +A_{p}y_{{t-p}}+e_{t},\,$$

Lorem ipsum dolor sit amet, consectetur adipiscing elit. Donec egestas justo in nulla bibendum ultrices. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Donec porttitor interdum purus, vel commodo lorem lacinia at. Suspendisse posuere ex non risus dapibus, vitae ullamcorper ex sollicitudin. Vestibulum interdum augue sit amet tortor mollis, vitae ornare diam pulvinar. Proin sapien elit, tempor non consectetur ut, mollis quis nunc.

## Granger Causality Test

Lorem ipsum dolor sit amet, consectetur adipiscing elit. Donec egestas justo in nulla bibendum ultrices. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Donec porttitor interdum purus, vel commodo lorem lacinia at. Suspendisse posuere ex non risus dapibus, vitae ullamcorper ex sollicitudin. Vestibulum interdum augue sit amet tortor mollis, vitae ornare diam pulvinar. Proin sapien elit, tempor non consectetur ut, mollis quis nunc. Maecenas tincidunt velit dolor, eu porta lacus ornare et. Ut in cursus felis. Etiam porta, ipsum sit amet pulvinar consectetur, nulla metus vehicula quam, ac blandit lacus ex et ante. Aliquam lacinia sagittis lobortis. Integer eros felis, gravida nec augue ut, blandit ultricies ipsum. Morbi aliquet at magna rhoncus vehicula. Suspendisse aliquam risus nunc, vitae eleifend lacus venenatis ut. Sed nisi eros, malesuada nec purus ut, tristique congue dui. Maecenas velit massa, fringilla vel tristique vitae, condimentum ut purus. Proin commodo eros iaculis, faucibus dui et, vehicula felis. Integer ligula urna, feugiat quis tristique vel, cursus nec quam. Mauris auctor sodales mauris, sed tristique quam sagittis in. Mauris ut bibendum enim. Maecenas consequat elit ut posuere aliquet. In hac habitasse platea dictumst. Integer est arcu, semper ac dolor at, pulvinar vehicula mi.

<!-- \newpage -->

@@

Lorem ipsum dolor sit amet, consectetur adipiscing elit. Donec egestas justo in nulla bibendum ultrices. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Donec porttitor interdum purus, vel commodo lorem lacinia at. Suspendisse posuere ex non risus dapibus, vitae ullamcorper ex sollicitudin. Vestibulum interdum augue sit amet tortor mollis, vitae ornare diam pulvinar. Proin sapien elit, tempor non consectetur ut, mollis quis nunc. Maecenas tincidunt velit dolor, eu porta lacus ornare et. Ut in cursus felis. Etiam porta, ipsum sit amet pulvinar consectetur, nulla metus vehicula quam, ac blandit lacus ex et ante. Aliquam lacinia sagittis lobortis. Integer eros felis, gravida nec augue ut, blandit ultricies ipsum. Morbi aliquet at magna rhoncus vehicula. Suspendisse aliquam risus nunc, vitae eleifend lacus venenatis ut. Sed nisi eros, malesuada nec purus ut, tristique congue dui. Maecenas velit massa, fringilla vel tristique vitae, condimentum ut purus. Proin commodo eros iaculis, faucibus dui et, vehicula felis. Integer ligula urna, feugiat quis tristique vel, cursus nec quam. Mauris auctor sodales mauris, sed tristique quam sagittis in. Mauris ut bibendum enim. Maecenas consequat elit ut posuere aliquet. In hac habitasse platea dictumst. Integer est arcu, semper ac dolor at, pulvinar vehicula mi.

@@

Pellentesque id purus eros. Curabitur rutrum ut nisl vel facilisis. Maecenas dapibus vel justo ut aliquet. Suspendisse at auctor sem. Pellentesque sit amet tortor vel arcu sollicitudin lacinia. Pellentesque condimentum, metus eget maximus aliquam, neque nisl.
Lorem ipsum dolor sit amet, consectetur adipiscing elit. Mauris ultricies felis in neque hendrerit semper. Sed quis ullamcorper nulla. Donec a mauris ut ipsum interdum tempor a et urna. Ut vehicula ligula sit amet nunc sodales, et facilisis felis vestibulum. Ut ullamcorper erat lorem, id congue ex ultricies ac. Nunc fermentum neque orci. Quisque congue, ligula eget tincidunt condimentum, nulla leo suscipit nibh, eu auctor tellus augue nec turpis. Nulla lorem dui, scelerisque non molestie et, pharetra in enim. Curabitur placerat leo a purus pellentesque, lobortis hendrerit justo posuere. Fusce condimentum porttitor purus, eget ullamcorper elit bibendum et. Cras quam mauris, mollis a cursus non, cursus id neque. Cras eget elit posuere, ornare dui non, posuere neque.

Ut tempor vestibulum augue sit amet hendrerit. Morbi rhoncus dignissim velit. Aliquam erat volutpat. Maecenas semper nulla nec sem tempus, eget elementum odio pulvinar. Donec sit amet nibh in risus dictum venenatis. Mauris non volutpat risus, vitae ornare dui. Praesent blandit leo nec dolor convallis, ut convallis lacus bibendum. Donec nec luctus quam. Nam congue suscipit ex, sed suscipit ligula vestibulum vel. Orci varius natoque penatibus et magnis dis parturient montes, nascetur ridiculus mus. Duis suscipit sem at enim commodo eleifend. Aenean eu malesuada justo. Sed finibus tincidunt lectus, eget facilisis leo porttitor id.

Morbi pellentesque ligula sed orci lacinia, in vulputate velit vestibulum. Nam commodo lectus odio, nec hendrerit purus consequat et. Sed vel iaculis ex. Maecenas mollis nisi id est pharetra facilisis. Fusce et tempor lorem. Etiam at facilisis nulla. Praesent sed erat odio. Integer sed dictum tellus, in finibus odio. Duis et rhoncus justo.
Pellentesque id purus eros. Curabitur rutrum ut nisl vel facilisis. Maecenas dapibus vel justo ut aliquet. Suspendisse at auctor sem. Pellentesque sit amet tortor vel arcu sollicitudin lacinia. Pellentesque condimentum, metus eget maximus aliquam, neque nisl.

Lorem ipsum dolor sit amet, consectetur adipiscing elit. Mauris ultricies felis in neque hendrerit semper. Sed quis ullamcorper nulla. Donec a mauris ut ipsum interdum tempor a et urna. Ut vehicula ligula sit amet nunc sodales, et facilisis felis vestibulum. Ut ullamcorper erat lorem, id congue ex ultricies ac. Nunc fermentum neque orci. Quisque congue, ligula eget tincidunt condimentum, nulla leo suscipit nibh, eu auctor tellus augue nec turpis. Nulla lorem dui, scelerisque non molestie et, pharetra in enim. Curabitur placerat leo a purus pellentesque, lobortis hendrerit justo posuere. Fusce condimentum porttitor purus, eget ullamcorper elit bibendum et. Cras quam mauris, mollis a cursus non, cursus id neque. Cras eget elit posuere, ornare dui non, posuere neque.

Ut tempor vestibulum augue sit amet hendrerit. Morbi rhoncus dignissim velit. Aliquam erat volutpat. Maecenas semper nulla nec sem tempus, eget elementum odio pulvinar. Donec sit amet nibh in risus dictum venenatis. Mauris non volutpat risus, vitae ornare dui. Praesent blandit leo nec dolor convallis, ut convallis lacus bibendum. Donec nec luctus quam. Nam congue suscipit ex, sed suscipit ligula vestibulum vel. Orci varius natoque penatibus et magnis dis parturient montes, nascetur ridiculus mus. Duis suscipit sem at enim commodo eleifend. Aenean eu malesuada justo. Sed finibus tincidunt lectus, eget facilisis leo porttitor id.

Morbi pellentesque ligula sed orci lacinia, in vulputate velit vestibulum. Nam commodo lectus odio, nec hendrerit purus consequat et. Sed vel iaculis ex. Maecenas mollis nisi id est pharetra facilisis. Fusce et tempor lorem. Etiam at facilisis nulla. Praesent sed erat odio. Integer sed dictum tellus, in finibus odio. Duis et rhoncus justo.
Pellentesque id purus eros. Curabitur rutrum ut nisl vel facilisis. Maecenas dapibus vel justo ut aliquet. Suspendisse at auctor sem. Pellentesque sit amet tortor vel arcu sollicitudin lacinia. Pellentesque condimentum, metus eget maximus aliquam, neque nisl.

Lorem ipsum dolor sit amet, consectetur adipiscing elit. Mauris ultricies felis in neque hendrerit semper. Sed quis ullamcorper nulla. Donec a mauris ut ipsum interdum tempor a et urna. Ut vehicula ligula sit amet nunc sodales, et facilisis felis vestibulum. Ut ullamcorper erat lorem, id congue ex ultricies ac. Nunc fermentum neque orci. Quisque congue, ligula eget tincidunt condimentum, nulla leo suscipit nibh, eu auctor tellus augue nec turpis. Nulla lorem dui, scelerisque non molestie et, pharetra in enim. Curabitur placerat leo a purus pellentesque, lobortis hendrerit justo posuere. Fusce condimentum porttitor purus, eget ullamcorper elit bibendum et. Cras quam mauris, mollis a cursus non, cursus id neque. Cras eget elit posuere, ornare dui non, posuere neque.

Ut tempor vestibulum augue sit amet hendrerit. Morbi rhoncus dignissim velit. Aliquam erat volutpat. Maecenas semper nulla nec sem tempus, eget elementum odio pulvinar. Donec sit amet nibh in risus dictum venenatis. Mauris non volutpat risus, vitae ornare dui. Praesent blandit leo nec dolor convallis, ut convallis lacus bibendum. Donec nec luctus quam. Nam congue suscipit ex, sed suscipit ligula vestibulum vel. Orci varius natoque penatibus et magnis dis parturient montes, nascetur ridiculus mus. Duis suscipit sem at enim commodo eleifend. Aenean eu malesuada justo. Sed finibus tincidunt lectus, eget facilisis leo porttitor id.

Morbi pellentesque ligula sed orci lacinia, in vulputate velit vestibulum. Nam commodo lectus odio, nec hendrerit purus consequat et. Sed vel iaculis ex. Maecenas mollis nisi id est pharetra facilisis. Fusce et tempor lorem. Etiam at facilisis nulla. Praesent sed erat odio. Integer sed dictum tellus, in finibus odio. Duis et rhoncus justo.Pellentesque id purus eros. Curabitur rutrum ut nisl vel facilisis. Maecenas dapibus vel justo ut aliquet. Suspendisse at auctor sem. Pellentesque sit amet tortor vel arcu sollicitudin lacinia. Pellentesque condimentum, metus eget maximus aliquam, neque nisl.

Morbi pellentesque ligula sed orci lacinia, in vulputate velit vestibulum. Nam commodo lectus odio, nec hendrerit purus consequat et. Sed vel iaculis ex. Maecenas mollis nisi id est pharetra facilisis. Fusce et tempor lorem. Etiam at facilisis nulla. Praesent sed erat odio. Integer sed dictum tellus, in finibus odio. Duis et rhoncus justo.Pellentesque id purus eros. Curabitur rutrum ut nisl vel facilisis. Maecenas dapibus vel justo ut aliquet. Suspendisse at auctor sem. Pellentesque sit amet tortor vel arcu sollicitudin lacinia. Pellentesque condimentum, metus eget maximus aliquam, neque nisl.

Morbi pellentesque ligula sed orci lacinia, in vulputate velit vestibulum. Nam commodo lectus odio, nec hendrerit purus consequat et. Sed vel iaculis ex. Maecenas mollis nisi id est pharetra facilisis. Fusce et tempor lorem. Etiam at facilisis nulla. Praesent sed erat odio. Integer sed dictum tellus, in finibus odio. Duis et rhoncus justo.Pellentesque id purus eros. Curabitur rutrum ut nisl vel facilisis. Maecenas dapibus vel justo ut aliquet. Suspendisse at auctor sem. Pellentesque sit amet tortor vel arcu sollicitudin lacinia. Pellentesque condimentum, metus eget maximus aliquam, neque nisl.

>>>>

The key motivation for our work is previous work showing that certain crimes like domestic violence saw a significantly smaller drop as compared to other crimes after the onset of the pandemic [[@boman2020has]], indicating the importance of the nature of the actual crime committed. Our focus on individual city data is due to known data collection issues for such research [@boman2020has] and is motivated by preliminary studies showing that the crime rates for specific crimes are largely dependent on the city [@ashby2020initial]. There are also other factors such as the George Floyd protests in June 2020 that we hypothesize may have resulted in a noticeable increase in the vandalism and theft crimes in certain cities.

Overall, our paper is a small contribution building upon and motivated by several recent papers that have covered the impact of the pandemic on the human psyche and crime rates across different states and countries [@boman2021global; @campedelli2020disentangling; @nivette2021global] alongside work prompting further research in the area [@eisner2020violence]. We achieve our goal of analyzing the pandemic data in conjunction with crime data obtained from the official city websites, using R statistical programming [@citeR], through time series modeling and also exploratory data analysis.

@@

In this study we employ Vector autoregressive models to model the relationship between each crime and COVID-19 cases, which are then, in turn, used for Granger causality tests and finally impulse response modelling.

@@

Vector autoregression (VAR) is a statistical model used to capture the relationship between multiple quantities as they change over a period of time. The main idea behind this model is to simply relate the present value for any variable at time $t$ to the past values of the variable. For our study, and each city, we fit a  VAR model between a given arbitrary crime denoted by $\text{CRIME}$ and the COVID-19 cases denoted by $\text{CASES}$.  A pth-order VAR is fitted where p is decided based on a simpler generalised additive model. The equation then takes the following form,

$$\text{CRIME}\displaystyle_{t}=c+A_{11}\text{CASES}_{t-1}+A_{12}\text{CASES}_{t-2}+\cdots +A_{1p}y_{t-p}+e_{1t},  $$

$$\text{CASES}\displaystyle_{t}=c+A_{21}\text{CRIME}_{t-1}+A_{22}\text{CRIME}_{t-2}+\cdots +A_{2p}y_{t-p}+e_{2t}$$

Here $A_i1 \ldots A_ip$ denote the coefficients for both the equations with $i\in \{1,2\}$ and $e_{it}$ denotes the error term. We fit these VAR models on the top five crimes for each city and then run Granger causality tests which are outlined below to establish causation.

\bigskip

## Granger Causality Test
 To determine if COVID-19 cases in a specific locality are useful in forecasting a specific crime we employ Granger causality testing. The null-hypothesis for our specific test is simply that an arbitrary crime $\text{CRIME}$ is not caused by the the COVID-19 cases denoted by $\text{CASES}$. The test employs the formerly fitted VAR model to compute a $p$ value with the test-statistic here following a $F$ distribution. It should be noted here that this tests only for direct causality and not indirect causality which may occur due to a third unknown variable.

Furthermore, we also conduct a simultaneous test for instantaneous Granger causality which is essentially a test that checks if both $x$ effects $y$ and $y$ effects $x$. The null hypothesis in this case is essentially that either  $x$ doesn't effect $y$ or $y$ doesn't effect $x$. More details about this test including the mathematical formulation are outlined in a much better fashion in a manuscript by @seth2007granger. 
<!-- \newpage -->

\bigskip

@@

Finally, we also employ impulse response modelling that builds upon Granger tests to show us how a change of one standard deviation in $\text{CASES}$ does to the rate of  $\text{CRIME}$ in the future. Impulse response modelling is a simple process to investigate such relationships, specifically ones which are significant on the Granger test. We employ impulse response modelling to visualise these relationships and demonstrate how due to a single new COVID-19 case, specific crimes may fluctuate in the future. Further detail on Impulse response modelling has been outlined in @koop1996impulse.

Next we investigate the criminal activity in each city using these modelling approaches and visualise the significant crimes in the subsequent section.

@@

## Overview of Results

During the COVID-19 pandemic, the world has seen several different changes both negative and positive. With increasing stay-at-home restrictions and various other policy measures, we aim to examine the impact that the pandemic has on criminal activity across several different cities in the United States including Chicago, Los Angeles, Seattle, and Philadelphia. The main goal of this study is to take a closer look at how different crimes varied across the pandemic as well as compare them to trends prior to the pandemic.

Consistent with early research that anticipated an increase in thefts and burglaries with the onset of the pandemic [@miller2020crime], our study found that there was a strong correlation between the rate of COVID-19 cases and crimes such as thefts and burglaries, especially in cities such as Seattle and Philadelphia, as seen in Figures \@ref(fig:r41) and \@ref(fig:r32). There was also a significant relationship between vehicle thefts and COVID-19 cases in cities such as Philadelphia and Chicago. as seen in Figures \@ref(fig:r31) and \@ref(fig:r12). Overall, our exploratory analysis also indicated similar trends, with the uptick in thefts and burglaries possibly occurring due to the lack of guardianship of assets due to people staying home.  Furthermore, our study indicated that crimes such as vandalism also show some relation with COVID-19 cases when it comes to cities like L.A, as seen in Figure \@ref(fig:r21), alongside other crimes such as assault in cities like Chicago, as observed in Figure \@ref(fig:r11). Overall it is noted that crimes that are better committed individually such as assault, theft, and burglary stood out more as compared to crimes committed in groups like vandalism. This is consistent with other research on crime in the United States which arrived at similar conclusions [@boman2020has].

The impulse response graphs show us another perspective with respect to these crimes. These graphs show us the future effects of a new COVID-19 case on a specific criminal activity. Based on the Figures  \@ref(fig:r12),  \@ref(fig:r31),  \@ref(fig:r32) &  \@ref(fig:r41) it can be noted that on average across all of the cities examined there is a high likelihood of at least one vehicle theft or burglary occurring after a unit increase in cases. Furthermore, based on our exploratory analysis carried out in Section 2, there seems to be no significant change in the frequency of occurrence of several crimes across several cities, pointing to a more bleak reality as compared to prior studies indicating an overall crime reduction.

These results in a way bring into question the significance of the overall decline in crime as there seem to be crimes like thefts and vehicle thefts that seem to have gone up significantly when compared to historical data. Furthermore, it seems like if the pandemic worsens in the future, we might see more crimes that are simply caused due to lack of guardianship as pointed out earlier. However, before arriving at such concrete conclusions we must take into consideration some of the limitations of this study and also explore future directions.

## Ethical Concerns and Limitations

Across studies that have examined crime during the COVID-19 pandemic, the lack of sufficient data has been a common limitation [@boman2020has; @ashby2020initial; @boman2021global; @campedelli2020disentangling; @nivette2021global]. In our study too, we find that unification of several datasets derived from different police units and cities is hard, and arriving at a concrete common conclusion based on them is even harder.  This study looked at data collected, cleaned, and stored by different organizations across the United States impacted in different ways by both crime and criminal activity, making it important to acknowledge that the conclusions we arrive at might not be truly generalizable.

When it comes to our modeling approach, we note that the VAR models used are quite limited in their ability as the only variables involved in our models are counts of the specific crime and COVID-19 cases. These models should not be used to forecast, COVID-19 or Crime cases as there are several other factors at play that influence either of their occurrences. Another major limitation and concern are to note that the Granger causality does not determine true causality and is only one of the many metrics that can be used to determine the cause-effect relationship between variables. The Granger test fails when both the involved variables are simultaneously affected by a third unknown variable, thus giving the illusion of causality. We only use the Granger test in our study to narrow down relevant variables but future work can certainly employ other more sophisticated Bayesian Causality metrics like GARCH  [@chen2017bayesian]. Finally, there are also several biases like underreporting of certain types of crimes and racial bias that plague crime data [@hetey2018numbers; @buil2021accuracy] that have to be acknowledged as well. It should also be noted that, due to stay-at-home measures and lack of access to police stations or support, several crimes especially rape and sexual assault are even more underreported than usual limiting the scope of our research [@palermo2011undercounting].

## Future Directions

Finally, our study opens the door for more future work exploring the impact brought upon by the COVID-19 pandemic on crime and the human psyche. With  the now availability of more comprehensive datasets like the global crime data set compiled by @nivette2021global, such an analysis can be further extended to other countries. With more sophisticated modelling approaches such as machine learning inspired recurrent neural networks to make more accurate predictions, our predictive approaches can be made more expressive and further improved to arrive at better informed conclusions. A more comprehensive study can also include other crimes such as domestic violence and hate crimes against racial minorities alongside cybercrime which seems to have increased over the pandemic due to increased internet use.
